### 2014-03-27 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding multilevel results for m=2

Monte Carlo solvers are effectively a stochastic realization of
Richardson's iteration and therefore we will first look at the
performance ofRichardson's iteration as a smoother:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I}-\ve{M}^{-1}\ve{A})\ve{x}^k\:,
\end{equation}
where $k$ is the iteration index and which is equivalently a Jacobi
iteration when Jacobi preconditioning is
used. Figure~\ref{fig:richardson} gives infinity norm of the error in
the solution vector\footnote{The solution to the homogenous problem is
  zero and therefore $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as
a function of iteration for wave numbers of 1, 5, and 10 on a grid
with $G = 100$.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{richardson.png}
  \end{center}
  \caption{\textbf{Convergence of Richardson's iteration as a function
      of iteration for a grid of size $G = 100$.} \textit{Richardson's
      iteration performs better for larger wave numbers and therefore
      more oscillatory modes.}}
  \label{fig:richardson}
\end{figure}
Immediately we note that the larger the wave number the better
Richardson's iteration performs. Per the spectral analysis in
\cite{briggs_multigrid}, this iteration sequence perfoms better for
higher wave numbers as the eigenvalue spectrum spans a smaller space
then less oscillatory modes. This behavior motivates the multigrid
approach where moving a smooth mode to a coarser grid makes that mode
appear more oscillatory relative to that coarser grid, thus improving
convergence for that particular mode.

Next we perform the same calculations using the adjoint Monte Carlo
solver presented in \cite{evans_monte_2012}. Before doing this
however, we must first modify the linear problem as the Monte Carlo
solver is a direct method and therefore a homogenous problem with a
right hand side of $\ve{0}$ will yield no samples. Instead, we will
solve the residual problem:
\begin{equation}
  \ve{A}\ve{d} = \ve{r}\:,
  \label{eq:residual_problem}
\end{equation}
where the residual of the homogenous problem is:
\begin{equation}
  \ve{r} = -\ve{A}\ve{x}^0\:,
  \label{eq:homogenous_residual}
\end{equation}
and the solution is computed as:
\begin{equation}
  \ve{x} = \ve{x}^0 + \ve{d}\:.
  \label{eq:residual_solution}
\end{equation}
Forming the problem in this way lets us directly apply the adjoint
Monte Carlo method to the homogenous problem and then apply the
effective correction, $\ve{d}$, to the initial guess to give the
solution. This approach is also equivalent to performing a single
iteration of Halton's method \cite{halton_sequential_1962}. Using this
formulation we can again solve the model problem with wave numbers of
1, 5 and 10 on a grid of size $G = 100$ but this time we vary the
number of histories used to compute the solution instead of the number
of iterations. Figure~\ref{fig:adjoint_mc} gives the results of these
calculations.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{adjoint_mc.png}
  \end{center}
  \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
      function of sampled histories for a grid of size $G = 100$.}
    \textit{Adjoint Monte Carlo performs better for smaller wave
      numbers and therefore smoother modes.}}
  \label{fig:adjoint_mc}
\end{figure}
Surprisingly, the Monte Carlo method performs better for smooth modes
than more oscillatory modes\footnote{This behavior was also observed
  for the forward Monte Carlo method presented in
  \cite{evans_monte_2012}.}. The results of these calculations are
counterintuitive given the fact that the Monte Carlo solver is
effectively a stochastic realization of Richardson's iteration and
therefore one should expect the same spectral behavior from the
results.

@@

%%---------------------------------------------------------------------------%%
\section{Results}
\label{sec:results}
To demonstrate the multilevel Monte Carlo algorithm we solve the
Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
with a grid of size $G = 1024$. For the coarsening parameter we will
first use a value of $M = 2$. In addition, we will introduce the
following standard figure of merit metric to assess the quality of the
Monte Carlo solver:
\begin{equation}
  FOM = \frac{1}{||\ve{e}||^2_{\infty} T}\:,
  \label{eq:figure_of_merit}
\end{equation}
where $||\ve{e}||_{\infty}$ is a measure of the variance of the
calculation and $T$ is the time required to solve the problem. The
larger the figure of merit, the more useful the solution scheme
is. For example, if two solutions schemes arrive at the same error but
the first calculation takes twice as long as the second, the figure of
merit will be twice as large for the second calculation.

>>>>

Monte Carlo solvers are effectively a stochastic realization of
Richardson's iteration and therefore we will first look at the
performance ofRichardson's iteration as a smoother:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I}-\ve{M}^{-1}\ve{A})\ve{x}^k\:,
\end{equation}
where $k$ is the iteration index and which is equivalently a Jacobi
iteration when Jacobi preconditioning is
used. Figure~\ref{fig:richardson} gives infinity norm of the error in
the solution vector\footnote{The solution to the homogenous problem is
  zero and therefore $||\ve{e}||_{\infty} = ||\ve{x}||_{\infty}$.} as
a function of iteration for wave numbers of 1, 5, and 10 on a grid
with $J = 100$.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{richardson.png}
  \end{center}
  \caption{\textbf{Convergence of Richardson's iteration as a function
      of iteration for a grid of size $J = 100$.} \textit{Richardson's
      iteration performs better for larger wave numbers and therefore
      more oscillatory modes.}}
  \label{fig:richardson}
\end{figure}
Immediately we note that the larger the wave number the better
Richardson's iteration performs. Per the spectral analysis in
\cite{briggs_multigrid}, this iteration sequence perfoms better for
higher wave numbers as the eigenvalue spectrum spans a smaller space
then less oscillatory modes. This behavior motivates the multigrid
approach where moving a smooth mode to a coarser grid makes that mode
appear more oscillatory relative to that coarser grid, thus improving
convergence for that particular mode.

Next we perform the same calculations using the adjoint Monte Carlo
solver presented in \cite{evans_monte_2012}. Before doing this
however, we must first modify the linear problem as the Monte Carlo
solver is a direct method and therefore a homogenous problem with a
right hand side of $\ve{0}$ will yield no samples. Instead, we will
solve the residual problem:
\begin{equation}
  \ve{A}\ve{d} = \ve{r}\:,
  \label{eq:residual_problem}
\end{equation}
where the residual of the homogenous problem is:
\begin{equation}
  \ve{r} = -\ve{A}\ve{x}^0\:,
  \label{eq:homogenous_residual}
\end{equation}
and the solution is computed as:
\begin{equation}
  \ve{x} = \ve{x}^0 + \ve{d}\:.
  \label{eq:residual_solution}
\end{equation}
Forming the problem in this way lets us directly apply the adjoint
Monte Carlo method to the homogenous problem and then apply the
effective correction, $\ve{d}$, to the initial guess to give the
solution. This approach is also equivalent to performing a single
iteration of Halton's method \cite{halton_sequential_1962}. Using this
formulation we can again solve the model problem with wave numbers of
1, 5 and 10 on a grid of size $J = 100$ but this time we vary the
number of histories used to compute the solution instead of the number
of iterations. Figure~\ref{fig:adjoint_mc} gives the results of these
calculations.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in]{adjoint_mc.png}
  \end{center}
  \caption{\textbf{Convergence of the adjoint Monte Carlo method as a
      function of sampled histories for a grid of size $J = 100$.}
    \textit{Adjoint Monte Carlo performs better for smaller wave
      numbers and therefore smoother modes.}}
  \label{fig:adjoint_mc}
\end{figure}
Surprisingly, the Monte Carlo method performs better for smooth modes
than more oscillatory modes\footnote{This behavior was also observed
  for the forward Monte Carlo method presented in
  \cite{evans_monte_2012}.}. The results of these calculations are
counterintuitive given the fact that the Monte Carlo solver is
effectively a stochastic realization of Richardson's iteration and
therefore one should expect the same spectral behavior from the
results.

@@

%%---------------------------------------------------------------------------%%
\section{Results}
\label{sec:results}
To demonstrate the multilevel Monte Carlo algorithm we solve the
Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
with a grid of size $J = 1024$. For the coarsening parameter we will
first use a value of $M = 2$. For simplicity, we will assume a basic
geometric multigrid formulation of the linear problem per
\cite{briggs_multigrid} with the the following prolongation operator
for $M=2$:
\begin{gather}
  \begin{subequations}
    x^h_{2j} = x_j^{2h} \\
    x^h_{2j+1} = \frac{1}{2} ( x_j^{2h} + x_{j+1}^{2h} )\:,
  \end{subequations}
  \label{eq:m2_prolongation}
\end{gather}
for $0 \leq j \leq \frac{J}{2} - 1$ and the restriction operator:
\begin{equation}
  x^{2h}_j = \frac{1}{4} ( x_{2j-1}^{h} + 2 x_{2j}^{2h} + x_{2j+1}^{2h} )
  \label{eq:m2_restriction}
\end{equation}
for $1 \leq j \leq \frac{J}{2} - 1$.

To measure performance of the algorithm, we will introduce the
following standard figure of merit metric to assess the quality of the
Monte Carlo solver:
\begin{equation}
  FOM = \frac{1}{||\ve{e}||^2_{\infty} T}\:,
  \label{eq:figure_of_merit}
\end{equation}
where $||\ve{e}||_{\infty}$ is a measure of the variance of the
calculation and $T$ is the time required to solve the problem. The
larger the figure of merit, the more useful the solution scheme
is. For example, if two solutions schemes arrive at the same error but
the first calculation takes twice as long as the second, the figure of
merit will be twice as large for the second calculation.
